### 2020-07-13 paper.tex
finish commenting out
Typos

See Section~\ref{sec:radon-nikodym-special-case} for a derivation of the Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d \mu$ for a special case, when the involution $f$ can be factored into (i) an involution on a countable partition of $Z$ and (ii) a family of continuously differentiable bijection on Euclidean spaces.

@@

The probabilistic programming language used in this paper is based on the dynamic modeling language in the Gen probabilistic programming system~\citep{gen-pldi}.
In this language, user tags each random choice with an explicit address in the code, as part of a \emph{random-choice expression} of the following form:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
{<address>} ~ <distribution>
\end{lstlisting}}
\end{tabular}
\end{center}
The left-hand side of the expression (green) specifies the address of the choice, and the right-hand side specifies the probability distribution to be sampled from.
For example, the following expression samples a non-negative integer from a geometric distribution and the value gets recorded in the trace at the address \texttt{"a"}:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
{"a"} ~ geometric(0.5)
\end{lstlisting}}
\end{tabular}
\end{center}
In addition to being recorded in the trace, the sampled value is returned to the running program.
That is, the random choice expression evaluates to the sampled value.
For example, the expression below evaluates to a an even non-negative integer:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
({"a"} ~ geometric(0.5)) * 2
\end{lstlisting}}
\end{tabular}
\end{center}
Probabilistic programs are typically constructed by combining multiple random choice expressions with regular Julia code (Gen is embedded in the Julia programming language~\citep{julia}).
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
a = ({"a"} ~ geometric(0.5))
{"b"} ~ bernoulli(a / (a + 1))
\end{lstlisting}}
\end{tabular}
\end{center}
This probabilistic program has an unbounded (countably infinite) number of possible traces it could produce.
One possible trace samples the value 1 at $\texttt{"a"}$ and the Boolean value $\false$ at \texttt{"b"}, and we denote this trace by $\{\mathtt{"a"} \mapsto 1, \mathtt{"b"} \mapsto \false\}$.
Note that the variable to which the first choice was assigned happens to be the same as the address.
This is not required, and often the random choice has no variable name in the program, as in the second random choice with address \texttt{"b"}.
But to reduce possibe unecessary typing, the following optional syntactic sugar will simultaneously assign the value to the program variable $\mathtt{a}$ and record it at trace address $\mathtt{a}$.\footnote{
More precisely, the address is the Julia symbol (interned string) \texttt{a}, constructed with Julia syntax \texttt{:a}.}
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
a ~ geometric(0.5)
\end{lstlisting}}
\end{tabular}
\end{center}
Probabilistic programs can make use of familiar control flow constructs, like loops and branching.
Stochastic control flow is also possible:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
i = 1
while ({i} ~ bernoulli(0.5))
    i = i + 1
end
\end{lstlisting}}
\end{tabular}
\end{center}
Note that for the probabilistic program to be valid, we require that every address is used at most once within every possible execution.

@@

\subsection{Handling Continuous Distributions}
To handle programs that sample continuous probability distributions the notion of a probability mass function on traces needs to be generalized to a probability density function on traces with respect to a certain reference measure.
The probabilistic program still defines a probability distibution (i.e. measure) on traces, but viavia  a density instead of a probability mass function.

@@

The measure-theoretic formalism above is capable of expressing probabilistic programs that use continuous distributions by setting the reference measure space $(V_a, \Sigma_a, \mu_a)$ for some addresses $a$ to use the Lebesgue-measure.
For example, consider the following probabilistic program, which encodes a generative model for data generated from a polynomial with unknown degree and unknown coefficients:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@gen function polynomials(x_coords)
    degree ~ |geometric(0.5)|
    noise ~ |inv_gamma(1, 1)|
    coeffs = [ ({("c", i)} ~ |normal(0, 2)|)
               for i in 0:degree ]
    for i=1:length(x_coords)
        x = x_coords[i]
        mu = coeffs' * x.^(0:length(coeffs)-1)
        {("y", i)} ~ |normal(mu, noise + 0.1)|
    end
end
\end{lstlisting}}
\end{tabular}
\end{center}
The addresses used by this generative function are:
\[
\mathcal{A} := \begin{array}{c}
\{\mathtt{:}\mathtt{noise}, \mathtt{:}\mathtt{degree}\} \bigcup
\{(\mathtt{"c"}, i) : i \in \integers_{\ge 0}\} \bigcup\\
\{(\mathtt{"y"}, i) : i \in \integers_{> 0}\}
\end{array}
\]
(Note that $\mathcal{A}$ could contain additional addresses that the program never samples.)
The random choice at address $\mathtt{:}\mathtt{degree}$ is discrete with $V_a = \integers_{\ge 0}$ and the other random choices are continuous.
Address $\mathtt{:}\mathtt{:noise}$ has $V_a = \reals_{> 0}$ and $(\mathtt{"c"}, i)$ and $(\mathtt{"y"}, i)$ have $V_a = \reals$, and all of these use the Lebesgue measure. % TODO double-check this
The density with respect to the reference measure $\mu$ is:
\begin{align*}
p(x) =
\begin{array}{l}
p_{\mathrm{geometric}(0.5)}(x[\mathtt{:}\mathtt{degree}]) \cdot\\
p_{\mathrm{inversegamma}(1, 1)}(x[\mathtt{:}\mathtt{noise}]) \cdot\\
\prod_{0=1}^{x[\mathtt{:}\mathtt{degree}]} p_{\mathrm{normal}(0,1)}(x[(\mathtt{"c"}, i)])\cdot\\
\prod_{i=1}^n p_{\mathrm{normal}(m(t, i), 0.1 + x[\mathtt{noise}])}(x[(\mathtt{"y"}, i)])
\end{array}
\end{align*}
where $m(t, i) := \sum_{j=0}^{x[\mathtt{:}\mathtt{degree}]} \mathtt{xcoords}[i]^j \cdot x[(\mathtt{"c"}, j)]$.

@@

@tracefn h from (tr1, tr2) to (tr3, tr4)
   is_birth = @read(tr2, :is_birth, :discrete)
   @write(tr4, :is_birth, !is_birth, :discrete)
   k = @read(tr1, :k, :discrete)
   weights = @read(tr1, :weights, :continuous)
   if is_birth
     @write(tr3, :k, k+1, :discrete)
     new_mu = @read(tr2, :new_mu, :continuous)
     new_var = @read(tr2, :new_var, :continuous)
     @write(tr3, (:mu, k+1), new_mu, :continuous)
     @write(tr3, (:var, k+1), new_var, :continuous)
     new_weights = add_weight(weights)
     @write(tr4, :deletion_idx, k+1, :discrete)
   else
     idx = @read(tr2, :deletion_idx, :discrete)
     @copy(tr1, (:mu, idx), tr4, :new_mu)
     @copy(tr1, (:var, idx), tr4, :new_var)
     for i in (idx+1):k
       @copy(tr1, (:mu, i), tr3, (:mu, i-1))
       @copy(tr1, (:var, i), tr3, (:var, i-1))
     end
     @write(tr3, :k, k-1, :discrete)
     new_weights = delete_weight(weights, i)
   end
   @write(tr3, :weights, new_weights, :continuous)
end
\end{lstlisting}}
\end{tabular}
\end{center}

>>>>

See Section~\ref{sec:radon-nikodym-special-case} for a derivation of the Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d \mu$ for a special case, when the involution $f$ can be factored into (i) an involution on a countable partition of $Z$ and (ii) a family of continuously differentiable bijections on Euclidean spaces.

@@

The probabilistic programming language used in this paper is 
% TODO: why "based on"?
based on 
the dynamic modeling language in the Gen probabilistic programming system~\citep{gen-pldi}.
In this language, user tags each random choice with an explicit address in the code, as part of a \emph{random-choice expression} of the following form:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
{<address>} ~ <distribution>
\end{lstlisting}}
\end{tabular}
\end{center}
The left-hand side of the expression (green) specifies the address of the choice, and the right-hand side specifies the probability distribution to be sampled from.
For example, the following expression samples a non-negative integer from a geometric distribution and the value gets recorded in the trace at the address \texttt{"a"}:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
{"a"} ~ geometric(0.5)
\end{lstlisting}}
\end{tabular}
\end{center}
In addition to being recorded in the trace, the sampled value is returned to the running program.
That is, the random choice expression evaluates to the sampled value.
For example, the expression below evaluates to a an even non-negative integer:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
({"a"} ~ geometric(0.5)) * 2
\end{lstlisting}}
\end{tabular}
\end{center}
Probabilistic programs are typically constructed by combining multiple random choice expressions with regular Julia code (Gen is embedded in the Julia programming language~\citep{julia}).
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
a = ({"a"} ~ geometric(0.5))
{"b"} ~ bernoulli(a / (a + 1))
\end{lstlisting}}
\end{tabular}
\end{center}
This probabilistic program has an unbounded (countably infinite) number of possible traces it could produce.
One possible trace samples the value 1 at $\texttt{"a"}$ and the Boolean value $\false$ at \texttt{"b"}, and we denote this trace by $\{\mathtt{"a"} \mapsto 1, \mathtt{"b"} \mapsto \false\}$.
Note that the variable to which the first choice was assigned happens to be the same as the address.
This is not required, and often the random choice has no variable name in the program, as in the second random choice with address \texttt{"b"}.
But to reduce possibe unecessary typing, the following optional syntactic sugar will simultaneously assign the value to the program variable $\mathtt{a}$ and record it at trace address $\mathtt{a}$.\footnote{
More precisely, the address is the Julia symbol (interned string) \texttt{a}, constructed with Julia syntax \texttt{:a}.}
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
a ~ geometric(0.5)
\end{lstlisting}}
\end{tabular}
\end{center}
Probabilistic programs can make use of familiar control flow constructs, like loops and branching.
Stochastic control flow is also possible:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
i = 1
while ({i} ~ bernoulli(0.5))
    i = i + 1
end
\end{lstlisting}}
\end{tabular}
\end{center}
Note that for the probabilistic program to be valid, we require that every address is used at most once within every possible execution.

@@

\subsection{Handling Continuous Distributions}
To handle programs that sample continuous probability distributions the notion of a probability mass function on traces needs to be generalized to a probability density function on traces with respect to a certain reference measure.
The probabilistic program still defines a probability distribution (i.e. measure) on traces, but via  a density instead of a probability mass function.

@@

The measure-theoretic formalism above is capable of expressing probabilistic programs that use continuous distributions by setting the reference measure space $(V_a, \Sigma_a, \mu_a)$ for some addresses $a$ to use the Lebesgue-measure.
For example, consider the following probabilistic program, which encodes a generative model for data generated from a polynomial with unknown degree and unknown coefficients:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@gen function polynomials(x_coords)
    degree ~ |geometric(0.5)|
    noise ~ |inv_gamma(1, 1)|
    coeffs = [ ({("c", i)} ~ |normal(0, 2)|)
               for i in 0:degree ]
    for i=1:length(x_coords)
        x = x_coords[i]
        mu = coeffs' * x.^(0:length(coeffs)-1)
        {("y", i)} ~ |normal(mu, noise + 0.1)|
    end
end
\end{lstlisting}}
\end{tabular}
\end{center}
The addresses used by this generative function are:
\[
\mathcal{A} := \begin{array}{c}
\{\mathtt{:}\mathtt{noise}, \mathtt{:}\mathtt{degree}\} \bigcup
\{(\mathtt{"c"}, i) : i \in \integers_{\ge 0}\} \bigcup\\
\{(\mathtt{"y"}, i) : i \in \integers_{> 0}\}
\end{array}
\]
(Note that $\mathcal{A}$ could contain additional addresses that the program never samples.)
The random choice at address $\mathtt{:}\mathtt{degree}$ is discrete with $V_a = \integers_{\ge 0}$ and the other random choices are continuous.
The address $\mathtt{:}\mathtt{noise}$ has $V_a = \reals_{> 0}$ and $(\mathtt{"c"}, i)$ and $(\mathtt{"y"}, i)$ have $V_a = \reals$, and all of these use the Lebesgue measure. % TODO double-check this
The density with respect to the reference measure $\mu$ is:
\begin{align*}
p(x) =
\begin{array}{l}
p_{\mathrm{geometric}(0.5)}(x[\mathtt{:}\mathtt{degree}]) \cdot\\
p_{\mathrm{inversegamma}(1, 1)}(x[\mathtt{:}\mathtt{noise}]) \cdot\\
\prod_{0=1}^{x[\mathtt{:}\mathtt{degree}]} p_{\mathrm{normal}(0,1)}(x[(\mathtt{"c"}, i)])\cdot\\
\prod_{i=1}^n p_{\mathrm{normal}(m(t, i), 0.1 + x[\mathtt{noise}])}(x[(\mathtt{"y"}, i)])
\end{array}
\end{align*}
where $m(t, i) := \sum_{j=0}^{x[\mathtt{:}\mathtt{degree}]} \mathtt{xcoords}[i]^j \cdot x[(\mathtt{"c"}, j)]$.

@@

@tracefn h from (tr1, tr2) to (tr3, tr4)
   is_birth = @read(tr2, :is_birth, :discrete)
   @write(tr4, :is_birth, !is_birth, :discrete)
   k = @read(tr1, :k, :discrete)
   weights = @read(tr1, :weights, :continuous)
   if is_birth
     @write(tr3, :k, k+1, :discrete)
     new_mu = @read(tr2, :new_mu, :continuous)
     new_var = @read(tr2, :new_var, :continuous)
     @write(tr3, (:mu, k+1), new_mu, :continuous)
     @write(tr3, (:var, k+1), new_var, :continuous)
     new_weights = add_weight(weights)
     @write(tr4, :deletion_idx, k+1, :discrete)
   else
     idx = @read(tr2, :deletion_idx, :discrete)
     @copy(tr1, (:mu, idx), tr4, :new_mu)
     @copy(tr1, (:var, idx), tr4, :new_var)
     for i in (idx+1):k
       @copy(tr1, (:mu, i), tr3, (:mu, i-1))
       @copy(tr1, (:var, i), tr3, (:var, i-1))
     end
     @write(tr3, :k, k-1, :discrete)
     new_weights = delete_weight(weights, idx)
   end
   @write(tr3, :weights, new_weights, :continuous)
end
\end{lstlisting}}
\end{tabular}
\end{center}
